The Evolution of Phenotypically Invariant Genetic Networks

### 2017-08-01 handnotes.tex
edited abstract and spelling

\begin{abstract}
Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantiative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

@@

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertaintity about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frusturate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

@@

  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we decribe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. In the neutral case, this set desribes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. Speciation typically occurs on timesclaes approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

Such a system $\mathcal{S}$, is composed of two equaitons,

@@

  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system. As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regultory element (or promoter) for gene $i$.

@@

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determiine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. (Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivlanet to its impulse response

@@

  The literature is filled with detailed observations of molecular systems and their diversity. There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. Despite a symmetry in functionality or phenotype these systems often differ, sometimes substationaly, at the molecular level. How many different mechanisms have the same function? We urge the reader to consider the metaphorical black box.

@@

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any valye}
    \end{figure}
 % \end{example}

@@

  \plr{amount of heritable variation}
  We quantify (roughly) the amount of heritable variation
  by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
  per change in concentration of $j$.
  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
  In the context of transcription factor networks this may be affected 
  not only by the binding strength of $j$ to the promoter region of $i$
  but also the effects of other transcription factors (e.g., cooperativity)
  and local chromatin accessibliity \citep{stefflova2013cooperativity}.
  For this reason, 
  the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
  (Recall that although these are best modeled through nonlinear terms,
  by linearizing we essentially consider first-order effects.)
  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
  in the presence of substantial molecular noise \citep{canalization,buffering},
  suggesting that substantial variation in the micro-scale dynamics we consider here
  may be necessary to produce relevant phenotypic effects downstream.
  \plr{replace "micro-scale" with sthg else or discuss earlier}
  Differences in $A_{ij}$ due to a sequence change are hard to measure 
  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
  or in expression levels (e.g., cis-eQTL).
  The former may be an overestimate, 
  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
  \citet{kasowski2010variation} found differential occupancy in 7.5\%
  of binding sites of a transcription factor (p65) between human individuals.
  \citet{verlaan2009targeted} showed that cis-regulatory variation
  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
  while \citep{lappalainen2013transcriptome} found that human population variation 
  explained about 3\% of expression variation.
  \plr{Get some data from at least one other species in here!}
  Taken together, this suggests that variation in the entries of $A$
  may be on the scale of 1\% between individuals of a population --
  doubtless varying substantially between species and between genes.

@@

  \plr{strength of selection}
  It seems certain that selection is not so strong that intrapopulation variation
  is strongly deleterious,
  so that if $u$ is the typical scale on which selection acts
  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
  then $u > \sigma$.
  However,
  a range of studies \plr{find them} have found evidence for weak stabilizing selection
  on regulatory SNPs and cis-eQTL.
  For instance, 
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
  This suggests that 
  the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
  so that perhaps $\sigma$ and $u$ are relatively close.
  This is as would be expected if available variation is held in check by mutation--selection balance
  rather than genetic drift.
  A conservative estimate would be that $u = 5 \sigma$;
  taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
  to effect a noticable drop in fitness.
  \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}

@@

  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficents as 
  (a) a ``cloud'' of width X about the population mean, 
  which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
  In fact, the population mean will not produce exactly the optimal phenotype,
  but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.

@@

        Thus if populations $1$ and $2$ are homogenous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
      \end{definition}
      \begin{figure}[H]
        \centering
        \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
        \caption{F1 hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
      \end{figure}
 %     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
 %       DMI examples...
 %       \begin{figure}[H]
 %         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
 %       \end{figure}
 %     \end{example}

@@

        \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$. If intrapopulation regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
    \end{example}
 %     \begin{example}[Not all Networks can Host Incompatibilities]
 %       convex sets cant have DMIs

@@

\section*{The rate of speciation under neutrality in allopatric populations}
   In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
   \begin{align*}
      \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
    \end{align*}
    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intrapopulation regulatory variance,
    \begin{align*}
      P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
    \end{align*}
    the population mean architecture will move at rate
    \begin{align*}
      \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
    \end{align*}
    where $T$ is time measured in generations. 
    USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
    Thus the squared difference of it's phenotype will be,
    \begin{align*}
      \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
    \end{align*}
    and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
       \begin{align*}
         \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
    \end{align*}
     It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
    \begin{align*}
      \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
      \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
    \end{align*}
    and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.

@@

  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). The organism may have a gene regulatory network $\mathcal{S}$ that can depoly a situation specific metabolic strategy. That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.

@@

  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architecutres, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.

@@

  Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. Here, we offer a tractable and simple model which includes these missing features. Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptualy (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}). The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.

  We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuissance. We have demonstrated a method to analytically determine the set of all phenotypically invarint gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities. We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.

  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intrapopulation variation and its effective population size. Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates. Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. This insight is a consequence of combinding our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypess quadratically, with evolutionary time. This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.

@@

        \item[(b)]
            \plr{conjecture:}
            The parameterization is unique
            if $P$ is futhermore chosen so that 
            each $P_x$ other than $P_\ro$ is a projection matrix,
            and that 
            \begin{align*}
                0
                =
                P_x^T P_y
            \end{align*}
            for all $(x,y)$ except $(\ro,\nrno)$.

>>>>

\begin{abstract}
Here we introduce an analytical theory to study the evolution of biological systems, such as gene regulatory networks. The evolutionary conservation of phenotype under selective and environmental stasis does not necessitate conservation of the underlying mechanism, as distinct molecular pathways can realize identical phenotypes. Here we give an exact expression for the set of all linear mechanisms with identical phenotypes, and expect evolution under neutrality to explore this set. We employ a quantitative genetic approach to model evolution under neutrality as a random process over the set of all phenotype-invariant mechanisms: only mutational tweaks to the pathway that leave the phenotype invariant are optimally fit. We show that there is never a unique linear system architecture for any phenotype and that the evolutionary exploration of these distinct and mutationally connected mechanisms can lead to the rapid accumulation of hybrid incompatibilities between allopatric populations and thus lead to the rapid formation of new species -- in fewer generations than there are breeding individuals in a population.
\end{abstract}

@@

In some fields that seek to fit parametric models to experimental data, such as control
theory, chemical engineering, and statistics, it is well known that mathematical models
can fundamentally be \emph{unidentifiable} and/or \emph{indistinguishable} -- meaning that 
there can be uncertainty about an inferred model's parameters or even its claims about
causal structure, even with access to complete and perfect data \citep{bellman1970structural, grewal1976identifiability, walter1984structural}. Models with different 
parameter schemes, or even different mechanics can be equally accurate, but still not
\emph{actually} agree with what is being modelled. In control theory, where electrical 
circuits and mechanical systems are often the focus, it is understood that there can be an 
infinite number of ``realizations'', or ways to reverse engineer the dynamics of a black box,
even if all possible input and output experiments on the black box are performed \citep{kalman1963mathematical, anderson1966equivalence, zadeh1976linear}. In chemical
engineering, those who study chemical reaction networks sometimes refer to the fundamental
unidentifiability of these networks as ``the fundamental dogma of chemical kinetics'' \citep{craciun2008identifiability}. In computer science, this is framed as the relationship among processes that simulate one another \citep{van2004equivalence}.
Although this may frustrate the occasional engineer or scientist, viewed from another angle,
the concepts of unidentifiability and indistinguishability can provide a starting point for
thinking about externally equivalent systems -- systems that evolution can explore, so long
as the parameters and structures can be realized biologically. In fact, evolutionary
biologists who study convergent versus parallel evolution, homology, and analogy are very familiar with such functional symmetries; macroscopically identical phenotypes in even very closely related species can in fact be divergent at the molecular and sequence level \citep{true2001developmental, tsong2006evolution, hare2008sepsid, vierstra2014mouse, stergachis2014conservation, taylor2016diverse, matsui2015regulatory}.

@@

  We derive an analytical description of the set of all linear biological systems with identical phenotypes -- that is we describe the set of all gene network architectures that yield identical phenotypes, and show that all biological systems can, in principal, can undergo systems drift. In the neutral case, this set describes a manifold that evolution explores leaving phenotype invariant with respect to mutation, and predicts that if two populations become reproductively isolated, hybrid incompatibility can occur, despite the absence of adaptation, directional selection, or environmental change. Speciation typically occurs on timescales approximately on the order of $N_{e}$ generations, where $N_{e}$ is the effective population size.

@@

Such a system $\mathcal{S}$, is composed of two equations,

@@

  where we sometimes refer to $h(t) := Ce^{A t}B$ as the \emph{impulse response} of the system. As $A$ specifies gene regulatory network architecture, the $i^\text{th}$ row of $A$ can be interpreted as the \emph{cis}-regulatory element (or promoter) for gene $i$.

@@

    If the rate each of these genes is expressed is a linear function of their concentrations, the dynamics of the system are given by
    \begin{align*}
      \dot{\kappa_{1}}(t) &= \kappa_{2}(t) + u(t) \\
      \dot{\kappa_{2}}(t) &= - \kappa_{1}(t) + u(t)
    \end{align*}
    where $\dot{\kappa}$ denotes the time derivative. The initial conditions $\kappa_{1}(0)$ and $\kappa_{2}(0)$, and the input $u(t)$ then determine the concentrations through time. If we record the regulatory coefficients in the matrix
    \begin{align*}
      A &= \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} ,
    \end{align*}
    and define the column vector $B = \begin{bmatrix} 1 & 1 \end{bmatrix}^{T}$, then in matrix notation the dynamics are
      \begin{align*}
      \dot{\kappa} &= A \kappa(t) + B u(t) .
      \end{align*}
      Since only the dynamics of gene-1 are directly relevant to biological function, the dynamics of interest are given by
      \begin{align*}
      \phi(t) &= C \kappa (t)
      \end{align*}
      where the row vector $C$ is defined as $C = \begin{bmatrix} 1 & 0 \end{bmatrix}$. (Note: if the dynamics of both genes were physiologically relevant to the cell cycle, we would set $C$ to be the identity matrix).
  %  \begin{align*}
  %    A = \begin{bmatrix} 0 & 1 \\ -1 & 0 \end{bmatrix} , \qquad B = \begin{bmatrix} 1 \\ 1 \end{bmatrix}, \qquad C = \begin{bmatrix} 1 & 0 \end{bmatrix}
  %  \end{align*}
 %       The oscillatory system $\Sigma$ is thus given as
 %   \begin{align*}
 %     \Sigma = \left \{ \begin{array}{ll} \dot{\kappa}(t) &= \begin{bmatrix} 
 %       0 & 1 \\ 
 %      -1 & 0 
 %       \end{bmatrix} \kappa(t) + \begin{bmatrix} 1 \\ 1 \end{bmatrix} u(t) \\ 
 %         \phi(t) &= \begin{bmatrix} 1 & 0 \end{bmatrix} \kappa(t) \end{array} \right.
 %    \end{align*}

      Since the input is simply an impulse, its phenotype is equivalent to its impulse response

@@

  The literature is filled with detailed observations of molecular systems and their diversity. There are examples of significant diversity in the networks underlying processes such as circadian rhythm \citep{sancar2008intelligent}, cell cycle control \citep{cross2011evolution, kearsey2003enigmatic}, pattern formation, and metabolism  \citep{lavoie2009rearrangements, martchenko2007transcriptional, dalal2016transcriptional, christensen2011unique, hartl2007induction, alam2013aspergillus}. Despite a symmetry in functionality or phenotype these systems often differ, sometimes substantially, at the molecular level. How many different mechanisms have the same function? We urge the reader to consider the metaphorical black box.

@@

\end{tikzpicture}
    \caption{A graphical depiction of the set of all externally equivalent cell cycle control networks, $A(\tau)$. $\tau$ can take any value}
    \end{figure}
 % \end{example}

@@

  \plr{amount of heritable variation}
  We quantify (roughly) the amount of heritable variation
  by $\sigma^2$, the genetic variance present in a population in a typical entry of $A$.
  The coefficient $A_{ij}$ measures how much the rate of net production of $i$ changes
  per change in concentration of $j$.
  It is generally thought that regulatory sequence change contributes much more to inter- and intraspecific variation
  than does coding sequence change affecting molecular structure \citep{schmidt2010vertebrate}.
  In the context of transcription factor networks this may be affected 
  not only by the binding strength of $j$ to the promoter region of $i$
  but also the effects of other transcription factors (e.g., cooperativity)
  and local chromatin accessibility \citep{stefflova2013cooperativity}.
  For this reason, 
  the mutational target size for variation in $A_{ij}$ may be much larger than the dozens of base pairs
  typically implicated in the handful of binding sites for transcription factor $j$ of a typical promoter region.
  (Recall that although these are best modeled through nonlinear terms,
  by linearizing we essentially consider first-order effects.)
  On the other hand, a diverse set of buffering mechanisms are thought to contribute to phenotypic stability
  in the presence of substantial molecular noise \citep{canalization,buffering},
  suggesting that substantial variation in the micro-scale dynamics we consider here
  may be necessary to produce relevant phenotypic effects downstream.
  \plr{replace "micro-scale" with sthg else or discuss earlier}
  Differences in $A_{ij}$ due to a sequence change are hard to measure 
  -- the closest available data we could find either related to variation in transcription factor binding site occupancy
  or in expression levels (e.g., cis-eQTL).
  The former may be an overestimate, 
  since it does not capture buffering effects (if for instance only one site of many needs be occupied for transcription to begin)
  and the latter probably measures changes in steady-state concentration (our $x_i$) rather than the rate of change.
  \citet{kasowski2010variation} found differential occupancy in 7.5\%
  of binding sites of a transcription factor (p65) between human individuals.
  \citet{verlaan2009targeted} showed that cis-regulatory variation
  accounts for around 2--6\% of expression variation in human blood-derived primary cells,
  while \citep{lappalainen2013transcriptome} found that human population variation 
  explained about 3\% of expression variation.
  \plr{Get some data from at least one other species in here!}
  Taken together, this suggests that variation in the entries of $A$
  may be on the scale of 1\% between individuals of a population --
  doubtless varying substantially between species and between genes.

@@

  \plr{strength of selection}
  It seems certain that selection is not so strong that intra-population variation
  is strongly deleterious,
  so that if $u$ is the typical scale on which selection acts
  -- e.g., a phenotype that differs from the optimum by $x$ has fitness of $\exp(-x^2/2u^2)$ --
  then $u > \sigma$.
  However,
  a range of studies \plr{find them} have found evidence for weak stabilizing selection
  on regulatory SNPs and cis-eQTL.
  For instance, 
  \citet{maria_and_sergey} (others?) found evidence that large-effect regulatory mutations are weakly selected against in Drosophila.
  This suggests that 
  the strength of selection on phenotype is sufficient to weakly constrain regulatory variation,
  so that perhaps $\sigma$ and $u$ are relatively close.
  This is as would be expected if available variation is held in check by mutation--selection balance
  rather than genetic drift.
  A conservative estimate would be that $u = 5 \sigma$;
  taking $\sigma=.01$ as above, this suggests that changes in phenotype of 5\% are sufficient
  to effect a noticeable drop in fitness.
  \plr{BUT $\sigma$ IS VARIATION IN $A$ NOT PHENOTYPE}

@@

  It therefore seems reasonable to coarsely model the time evolution of population variation in network coefficients as 
  (a) a ``cloud'' of width X about the population mean, 
  which (b) moves as an unbiased Brownian motion through the set of network coefficients that give the optimal phenotype.
  In fact, the population mean will not produce exactly the optimal phenotype,
  but it will be convenient to refer to this closest point on the optimal set as ``the population mean''.

@@

        Thus if populations $1$ and $2$ are homogeneous $A(0)$ and $A(2)$, respectively, we say that they are completely incompatible as $\mathcal{I} = 0$.
      \end{definition}
      \begin{figure}[H]
        \centering
        \includegraphics[width=0.5\textwidth, height=0.25\paperheight]{expF1}
        \caption{F1 hybrid (orange) and parental (blue) phenotypic oscillator dynamics for an $A(0)$ by $A(2)$ cross. The hybrid fails to oscillate and exhibits qualitatively different dynamics.}
      \end{figure}
 %     \begin{example}[F1 Reproductive Incompatibility in an Oscillating Gene Network]
 %       DMI examples...
 %       \begin{figure}[H]
 %         \includegraphics[width=0.5\textwidth]{A2_A3_osci_F1_hyb}
 %       \end{figure}
 %     \end{example}

@@

        \jss{$A(2)$ and $A(2.1)$ differ in regulatory interaction strengths by $5.09\%$. If intra-population regulatory variation is approximately $5\%$, then this level of divergence is expected after $\sqrt{\frac{t}{N_{e}} = 1}$ generations.}
    \end{example}
 %     \begin{example}[Not all Networks can Host Incompatibilities]
 %       convex sets cant have DMIs

@@

\section*{The rate of speciation under neutrality in allopatric populations}
   In general terms, if fitness $\mathcal{F}$ is some function $g$ of phenotypic divergence from an optimal scaled by the strength of selection $\beta$,
   \begin{align*}
      \mathcal{F} = g \left( \frac{\phi - \phi_{0}}{\beta} \right)
    \end{align*}
    And a population $P$ is normally distributed around a mean gene network architecture $A(\tau)$ and $\sigma^{2}$ is the intra-population regulatory variance,
    \begin{align*}
      P_{T} \sim \mathcal{N} \left(A(\tau), T \frac{\sigma^{2}}{N_{e}}\right)
    \end{align*}
    the population mean architecture will move at rate
    \begin{align*}
      \Delta \tau = \sigma \sqrt{\frac{T}{N_{e}}}
    \end{align*}
    where $T$ is time measured in generations. 
    USING GENERAL MATH FROM MULTIVARIATE TAYLOR SERIES.... we know that the phenotype of a first generation $F_{1}$ hybrid diverges from that of its parents (assuming both parents are at an optimum and phenotypically equivalent) quartically as a function of $\Delta \tau$. 
    Thus the squared difference of it's phenotype will be,
    \begin{align*}
      \Delta \Phi^{2}_{1} = c_{1}^{2} \sigma^{4} \left( \frac{T}{N_{e}} \right)^{2}
    \end{align*}
    and $F_{1}$ fitness will decline linearly with respect to $\frac{T}{N_{e}}$. ARGUMENT FOR WHY $\sigma^{2} c_{1} \sim \beta$ HERE.
       \begin{align*}
         \Delta \mathcal{F}_{1} = \left( \frac{\Delta \Phi_{1}}{\beta} = \frac{c_{1} \sigma^{2} \frac{T}{N_{e}}}{\beta} \approx \frac{T}{N_{e}} \right)
    \end{align*}
     It follows FROM THE SAME MATH ABOUT THE MULTIVAR TAYLOR that the phenotypes of second generation $F_{2}$ hybrid crosses diverges quadratically as a function of $\Delta \tau$, 
    \begin{align*}
      \Delta \Phi^{2}_{2} &= c_{2}^{2} \left( \sigma \sqrt{\frac{T}{N_{e}}} \right)^{2} = c_{2}^{2} \frac{T}{N_{e}} \\
      \Delta \mathcal{F}_{2} &= \left( \frac{\Delta \Phi_{2}}{\beta} = \frac{c_{2} \sigma \sqrt{\frac{T}{N_{e}}}}{\beta} \approx \sqrt{\frac{T}{N_{e}}} \right)
    \end{align*}
    and that $F_{2}$ fitness drops rapidly -- at approximately rate $\sqrt{\frac{T}{N_{e}}}$. ARGUMENT FOR WHY $c_{2} \sigma \sim \beta$ HERE.

@@

  Consider an organism that can metabolize two different sugars $s_{1}$ and $s_{2}$ (present at logarithmic concentrations $u_{1}$ and $u_{2}$ respectively), with enzymes $e_{1}$ and $e_{2}$ (with log concentration denoted as $\phi_{1}$ and $\phi_{2}$). Further suppose that one of the sugars $s_{2}$ is the preferred energy source (perhaps it contains significantly more energy than the other sugar or is otherwise more efficiently metabolized). The organism may have a gene regulatory network $\mathcal{S}$ that can deploy a situation specific metabolic strategy. That is depending on both $u_{2}$ and $u_{1}$ the organism will synthesize an appropriate $\phi_{1}$ and $\phi_{2}$. Furthermore, consider this system to contain at least two transcription factors, whose log concentrations are given by $\kappa_{1}$, $\kappa_{2}$, $\dots \kappa_{n}$.

@@

  We can see that changing coordinates on this system (with any invertible $V \in \mathbb{R}^{2 \times 2}$) will find new system architectures, as shown before, however, we can also apply the Kalman decomposition to find systems in different dimensions -- that is systems that employ more than $2$ transcription factors, yet have the same external dynamics/phenotype. This may happen if a system is co-opted from another, or may be a consequence of gene duplication and deletion.

@@

  Theoretical models in evolution and population genetics often lack the molecular details of physiology or of the genotype-phenotype map. Here, we offer a tractable and simple model which includes these missing features. Further, we provide, in clear mathematical language, an analytical description of phenomena hitherto only discussed verbally and conceptually (phenogenetic drift \citep{weiss2000phenogenetic}, developmental systems drift \citep{true2001developmental}, biological degeneracy \citep{edelman2001degeneracy}, \emph{etc.}). The tractability and relative simplicity of this exposition enables the interested biologist to work out by hand, if desired, the dynamics of a genetic system, as well as perturbations to the system -- an attribute not likely to be found in less tractable models and simulations.

  We have suggested an interpretation of system identification: to see it as an evolutionarily neutral manifold, and not simply a computational nuisance. We have demonstrated a method to analytically determine the set of all phenotypically invariant gene networks; by a simple change of coordinates in the minimal configuration, or more generally by applying the Kalman decomposition in higher dimensions. Further, we emphasize that evolution proceeds through this high dimensional space as stochastic coordinate transformation, constrained by sexual reproduction and selection. This set is explored over evolutionary time when phenotype is conserved, and can lead to a diverse set of consequences, including the accumulation of Dobzhansky-Muller incompatibilities. We emphasize that these incompatibilities are a consequence of recombining different, yet functionally equivalent, mechanisms.

  Furthermore, using a quantitative genetic approach, we estimated that a genetically variable population will drift in neutral system space at a rate determined by its intra-population variation and its effective population size. Because mechanistically distinct yet phenotypically equivalent biological systems can fail to produce viable hybrids, we predict allopatric populations to speciate at a rate on the order of $N_{e}$ under reasonable population genetic parameter estimates. Additionally we see second-generation hybrid fitness plummet much faster than that of first-generation hybrids. This insight is a consequence of combining our mechanistic model with a quantitative genetic one: we observe that $F_{1}$ phenotypes diverge quartically, and $F_{2}$ phenotypes quadratically, with evolutionary time. This result is also consistent with Haldane's rule; that if only one hybrid sex is inviable or sterile it is likely the heterogametic sex. The consistency comes from gene networks localized to the sex chromosomes functioning as an $F_{2}$ hybrid cross within a diploid $F_{1}$ heterogamete as there is only one sex chromosome.

@@

        \item[(b)]
            \plr{conjecture:}
            The parameterization is unique
            if $P$ is furthermore chosen so that 
            each $P_x$ other than $P_\ro$ is a projection matrix,
            and that 
            \begin{align*}
                0
                =
                P_x^T P_y
            \end{align*}
            for all $(x,y)$ except $(\ro,\nrno)$.
